---
title: "Intro to Numerical Data, Continuing Intro to R"
include_overview: true
week: 2
readings:
 - This weeks topics cover remaining parts of OIS 1.1, 1.2, 1.3, 2.1; OIS 1.4 & 2.3 (we won't cover 2.3 yet, but will cover the concepts presented here later in the class so feel free to read it over now).
 - We will also cover ITR 1, 2, 5, 6, 7, 12 (ITR can be skimmed or read in depth depending on comfort level with coding)
questions:
 - How can we analyze numerical data?
 - TBD
objectives:
 - TBD
keypoints:
 - "TBD"
source: Rmd
---


```{r setup, include=FALSE}
source("../bin/chunk-options.R")
source("../bin/functions.R")
```
<!-- JPN: attribution formatting -->
<style type="text/css" rel="stylesheet">
* {.attribution{
     position:absolute;
     bottom:0;
     right:0;
     font-size:0.5em
 }}
</style>

## Summary statistics and histograms with data in R

### Review

Last time we talked about some *summary statistics* and how to calculate them for data that is made up of numbers:
  * the *mean* is the sum of all data values divided by the number of data points
  * the *median* can be thought of as the "middle" number that splits the data into two halves if we order numerical data from smallest to largest
  * the *standard* deviaion is the square root of something called the *varience* which is roughly the average distnace of data values from the mean
  * the *first/third quartiles* delinate where 25% of the data falls below (first) and above (third) if we ordered the data from smallest to largest (the *median* is like the 50% or second quartile)

Let's do some analysis with "fake" class data showing programming histories of students in the class.

```{r d1p2}
classData = read.csv("data/formatted_class_answers.csv", stringsAsFactors=T)
# note: this is a placeholder for real data
```

Make sure this file is stored somewhere you can remember!  You can put it in the same directory as this file (or whatever R-script you are working from) or you can specify a location.  For example, on my Mac I can specify the default `Downloads` folder as the location with:

~~~
classData = read.csv("~/Downloads/formatted_class_answers.csv", stringsAsFactors=T)
~~~
{: .language-r}

What is in this dataset?

```{r d2p2}
classData
```